---
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(astsa)
library(tseries)
library(ggplot2)
library(dplyr)
library(lubridate)
library(rugarch)
library(zoo)
```

\begin{center}
\LARGE\bfseries
Modeling Microsoft Stock Price Over Time: Utilizing ARIMA and GARCH  
\end{center}

\begin{center}
\large
Will Markevitch \\
wmarkevitch@ucsb.edu \\
Perm No: 4253167 \\
June 13, 2024
\end{center}

## Project Abstract           

This project utilizes Microsoft stock price data (MSFT) from the beginning of 2016 to the end of 2020 to 
                                              
forecast its future value. This is done by utilizing time series techniques such as ARIMA to model the 
                                     
trend and GARCH to model the volatility. By analyzing the historical stock prices and applying these 
          
advanced statistical models, we aim to provide accurate forecasts and insights into the stock price movements 
                  
of Microsoft.
            
\newpage

## 1. Introduction         
             
The goal of this project is to analyze and forecast the stock price of Microsoft using time series modeling 
                              
techniques. The choice of Microsoft as the subject of this analysis is motivated by its significant influence 
                                
in the tech industry and its substantial stock market presence. Given the company's pivotal role in technology 
                             
and its robust business strategies, understanding its stock price behavior can provide valuable insights.            
        
                    
In the past, various studies have employed time series models to forecast stock prices. Techniques such as 
           
ARIMA (AutoRegressive Integrated Moving Average) have been widely used due to their capability to model 
            
the linear dependencies in the data. However, stock prices are also known for their volatility, which is 
           
better captured by models like GARCH (Generalized AutoRegressive Conditional Heteroskedasticity).              
             
                              
For this project, we will use ARIMA to model the overall trend and GARCH to capture the volatility in 
                
Microsoft's stock prices. The data set spans from April 2015 to April 2021, but for this analysis, we will 
             
focus on the period from January 2016 to December 2020, which covers significant events including the rise 
                  
of cloud computing and the impact of the COVID-19 pandemic.         
         
                               
The analysis will involve data preprocessing, model fitting, and diagnostics to ensure the models are 
              
well-specified. The key findings and insights will be discussed, and potential areas for future study will 
                      
be highlighted to improve the robustness and accuracy of the forecasts.        
           
\newpage

## 2. Data Description

The dataset used in this project spans from April 2015 to April 2021, covering Microsoft's daily share prices 
   
over a period of six years, with all values expressed in US dollars and non-negative. Comprising 1511 daily 
    
records, the dataset offers a detailed view of price trends and fluctuations. It was sourced from Google's
     
public records using the 'GOOGLEFINANCE' command in Google Sheets, ensuring the data's accuracy 
         
and reliability. It should be noted that for simplicity of analysis and due to the nature of stock data 
         
(no weekends/holidays), I opted to average data for each month and subset from January 1st 2016 to 
                 
December 31st 2020. This comprehensive dataset is essential for analyzing how Microsoft's stock prices 
             
evolved through the late 2010s and through the beginning of the global Covid-19 pandemic. The focus on 
           
Microsoft is due to its significant business strategies involving expansions and acquisitions, which 
            
directly impact its stock prices. Moreover, as a leading tech company, Microsoft's performance during 
          
the pandemic serves as a crucial indicator of the tech sector's resilience. With the future of AI already 
            
among us, I wanted to also look into the historical performance of one of the leading companies in AI 
                
products. By studying this dataset, we aim to understand the impact of Microsoft's strategic decisions 
                           
on their stock prices and gauge the overall health and trends of the technology sector during a critical 
                 
period. The dataset can be accessed on Kaggle via the following link: [Microsoft Stock Time Series Analysis](https://www.kaggle.com/datasets/vijayvvenkitesh/microsoft-stock-time-series-analysis).

\newpage

## 3. Methodology

### 3.1 SARIMA (p, d, q) x (P, D, Q) model 
Seasonal Auto Regressive Integrated Moving Average (SARIMA) is a time series forecasting method that 
                   
accounts for seasonality in the data it is provided. It has the capabilities of an ARIMA model, but 
                  
also incorporates seasonal elements utilizing season differencing and seasonal auto regressive and 
             
moving average terms. This makes it a good mixture of capturing short-term patterns, such as week to
         
week patterns or day to day patterns in the data, but also capturing longer term seasonality trends
         
like year over year changes in stock price for our analysis. The general SARIMA model is represented by: $$\text{SARIMA}(p, d, q) * (P, D, Q)_s$$
      
Where:                
- $p$: Non-seasonal autoregressive order       
- $d$: Non-seasonal differencing order       
- $q$: Non-seasonal moving average order      
- $P$: Seasonal autoregressive order      
- $D$: Seasonal differencing order     
- $Q$: Seasonal moving average order     
- $s$: Length of the seasonal cycle                  
           
Due to the sub-setting of data and potential misrepresentation of seasonal patterns, I opted to remove
         
the seasonality component and employ an Autoregressive Integrated Moving Average (ARIMA) model for
        
forecasting Microsoft's monthly stock prices. This decision ensures a more straightforward and potentially
           
more accurate modeling approach for the given dataset, which might not exhibit strong seasonal 
        
characteristics.

### 3.2 GARCH model
                        
Generalized AutoRegressive Conditional Heteroskedasticity (GARCH) is a time series model designed 
                                     
to estimate the volatility of stock prices. Unlike traditional ARIMA models that focus on the mean 
                                  
of the series, GARCH models are particularly effective in capturing and forecasting the variability 
                             
or volatility of the data over time. This makes GARCH models particularly useful for financial time 
                                    
series, where volatility clustering (periods of high volatility followed by periods of low volatility) is a 
          
common phenomenon.
                
The GARCH model is defined by two components: the autoregressive (AR) term and the moving average 
          
(MA) term, both of which are applied to the variance of the series rather than the series itself. The basic 
                     
GARCH(p, q) model can be expressed as: $$\sigma^2_t = \alpha_0 + \sum^{q}_{i=1}\alpha_i\epsilon^2_{t-i} + \sum^{p}_{j=1}\beta_j\sigma^2_{t-j}$$      
       
Where:     
        
- $\sigma_t^2$: The conditional variance at time t.       
- $\alpha_0$: A constant term.      
- $\alpha_i$: The coefficients for the lagged squared residuals (innovations).      
- $\epsilon_{t-i}^2$: The lagged squared residuals from the mean equation.       
- $\beta_j$: The coefficients for the lagged conditional variances.     
        
In our analysis, we used a GARCH(1, 1) model in conjunction with an ARIMA(0, 1, 0) model to capture 
                       
both the trend and the volatility of Microsoft's monthly stock prices. This combined approach allows 
                        
us to model the mean and variance of the series effectively, accounting for the persistence in volatility 
             
that is often observed in stock prices. In a future study, I could utilize this in the forecasting         
      
for a more accurate prediction.

\newpage

## 4. Results

### Arima Model Fitting              
         
**Note:** Remember due to the nature of stock data (no weekends/holidays), I opted to average data for each month and subset from January 1st 2016 to December 31st 2020.    
              
### Plotting the Original Data and ACF & PACF  
       
```{r, message = FALSE}
msft <- read.csv("Microsoft_Stock.csv", stringsAsFactors = FALSE)

msft$Date <- mdy_hms(msft$Date)

msft <- msft %>%
  filter(Date >= as.Date("2016-01-01"), Date <= as.Date("2020-12-31"))

msft_monthly <- msft %>%
  group_by(Year = year(Date), Month = month(Date)) %>%
  summarise(Average.Close = mean(Close))

msft_ts <- ts(msft_monthly$Average.Close, start = c(2016, 1), end = c(2020, 12), frequency = 12)

ts.plot(msft_ts, xlab = "Year", ylab = "Average Close Price (USD $)", main = "Microsoft Stock Monthly Average Close Price")

acf2(msft_ts)
```
       
We can see that the data has a clear upward trend, indicating that it is non-stationary, so we need to difference our data to try and make it stationary.
  
### Determining the Difference              
```{r}
msft_ts_diff <- diff(log(msft_ts))
ts.plot(msft_ts_diff, xlab = "Year", ylab = "Log Diff Average Close Price (USD $)", main = "Log Diff Microsoft Stock Monthly Average Close Price")
```
         
This looks much more stationary (outlier in early 2020 - covid pandemic), so we can check the ACF and PACF graphs for determining ARIMA models to test.   
    
```{r}
acf2(msft_ts_diff, 20)
```
             
From the ACF and PACF graphs, it isn't clearly decipherable which values for $p$ and $q$ should be used for the ARIMA modeling. So instead, I'll test various values and check for alignment with the qq normality plot and lowest AIC and BIC.
     
### Model Selection: Estimation Results and Diagnostics                            
```{r}
sarima(msft_ts, 0, 1, 0, 0, 0, 0, 0) # Non-seasonal ARIMA(0, 1, 0) with zero mean
sarima(msft_ts, 1, 1, 0, 0, 0, 0, 0) # Non-seasonal ARIMA(1, 1, 0) with zero mean
sarima(msft_ts, 0, 1, 1, 0, 0, 0, 0) # Non-seasonal ARIMA(0, 1, 1) with zero mean
sarima(msft_ts, 1, 1, 1, 0, 0, 0, 0) # Non-seasonal ARIMA(1, 1, 1) with zero mean
```
          
### Forecasting                       
From the analysis above, the model with the lowest AIC and BIC was a non-seasonal ARIMA(0, 0) model with a $d = 1$., so we can forecast the next 12 data points below:  
       
```{r}
sarima.for(msft_ts, n.ahead = 12, 0, 1, 0)
```

### GARCH Fitting           
        
### Testing Various GARCH Models            
         
As we found that our Microsoft data is best represented by an ARIMA(0, 1, 0) model, when calculating
        
the ideal GARCH model we can use the differenced data and ARMA(0, 0). Here I'll test GARCH(1, 1),
       
GARCH(2, 1) and GARCH(1, 2):
    
```{r}
msft_garch11 <- ugarchspec(variance.model=list(model="sGARCH",garchOrder=c(1,1)),
                mean.model=list(armaOrder=c(0,0)),distribution.model="norm")

msft_garch11_summary <- ugarchfit(data = msft_ts, spec = msft_garch11)

msft_garch11_summary

msft_garch21 <- ugarchspec(variance.model=list(model="sGARCH",garchOrder=c(2,1)),
                mean.model=list(armaOrder=c(0,0)),distribution.model="norm")

msft_garch21_summary <- ugarchfit(data = msft_ts, spec = msft_garch21)

msft_garch21_summary

msft_garch12 <- ugarchspec(variance.model=list(model="sGARCH",garchOrder=c(1,2)),
                mean.model=list(armaOrder=c(0,0)),distribution.model="norm")

msft_garch12_summary <- ugarchfit(data = msft_ts, spec = msft_garch12)

msft_garch12_summary
```
               
### Diagnostics            
        
Now that the models are defined we can evaluate their diagnostics as follows:
            
```{r}
plot(msft_garch11_summary, which = 'all')
plot(msft_garch21_summary, which = 'all')
plot(msft_garch12_summary, which = 'all')
```

### Model Selection           
         
Looking at our normality (QQ) plots and the Akaline (AIC) values, GARCH(1, 1) appears to be the best 
             
fit for our differenced data. 
               
\newpage

## 5. Conclusion and Future Study
         
### Conclusion     
         
From the monthly data, we determined that Microsoft's stock prices most closely follow an 
                           
ARIMA(0, 1, 0) model, combined with a GARCH(1, 1) model to capture volatility. The ARIMA 
                            
model helps us understand the trend and differencing required to make the series stationary, 
                               
while the GARCH model captures the volatility clustering commonly observed in financial time 
                  
series. Our results suggest that this combination is effective in modeling and forecasting 
                           
Microsoft stock prices on a monthly basis, providing valuable insights into future price movements.
        
### Future Study                      
             
For future improvements, several approaches can be taken to enhance the robustness and accuracy of 
              
the model:        
              
- Daily Data Usage:          
                    
Incorporating all available daily data, rather than aggregating to monthly data, would allow 
                                    
for a more granular analysis. Although this approach would involve more complex data 
                          
cleaning and setup, it could significantly improve model accuracy by capturing finer details and 
                      
more frequent seasonal patterns.
                 
- Seasonality Consideration: 

With the full set of 1511 daily records, we could better account for seasonality in the data.    
                   
This would allow for the potential use of a Seasonal ARIMA (SARIMA) model, which could capture  
                            
seasonal trends and provide a more nuanced understanding of stock price movements.       
               
- Alternative Models:            
          
Exploring alternative modeling techniques could provide new insights and potentially improve 
                                             
forecasting accuracy. For instance, spectral analysis methods could be employed to identify and 
                           
model periodicities in the data that are not captured by ARIMA and GARCH models. Additionally,          
                 
other advanced time series models, such as the State Space Model or Long Short-Term Memory 
                    
(LSTM) neural networks, could be considered for capturing complex patterns in stock prices.
           
- External Factors: 

Incorporating external variables, such as macroeconomic indicators, company financials, or industry 
                               
trends, could enhance the model by providing additional context and improving its predictive power.
                     
          
### By implementing these improvements, we can achieve a more detailed and accurate understanding of stock price behavior, leading to better forecasting and investment strategies.
           
\newpage

## 6. Appendix and Source Code
       
**All plots used throughout the project have the associated code in line with it.**